---
title: "Baseline Simulations"
author: "Or Duek & Tobias R. Spiller"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: yes
---

```{r}
# load libraries
require(ggplot2)
require(tidyverse)
require(lme4)
require(nlme)
require(lmerTest)
require(emmeans)
require(cowplot)
```

# Introduction
Here we will simulate randomized controlled trial of two groups (placebo and treatment) for 3 (or more) time points.
The main idea is to test the effect of using the baseline score as a covariate in a mixed level model.

# Generating Data
First, the total number of patients
```{r}
n = 100 # number of subjects
score <- rnorm(n, 55, 10) # generate data for PCL scores of PTSD patients
hist(score)
mean(score)
```

Now we shal divide them to two groups.
Group 1 (trt) will improve an average of 10 points at time 2 and another 5 at time 3
Group 2 (placebo) will improve an average of 2 points on time 2 and 1 on time 3

```{r}
scoreTrt <-  score[1:(n/2)]
scorePlacebo <- score[((n/2)+1):n]
treatment <- data.frame(score = scoreTrt)
placebo <- data.frame(score = scorePlacebo)
treatment$group <- 'trt'
placebo$group <- 'placebo'
# and combine into your new data frame vegLengths
both <- rbind(treatment, placebo)
ggplot(both, aes(x=score, fill = group)) + geom_density(alpha = 0.2) + theme_minimal()
```
So we see we start with similar distribution.
Now lets generate the change

```{r}
# in treatment group we remove an average of 10 points
changeTrt1 <- rnorm((n/2),10,5)
changeTrt2 <- rnorm((n/2),5,5)

treatment$time2 <- treatment$score - changeTrt1 
treatment$time3 <- treatment$time2 - changeTrt2 

# now for the plcebo group
changePlacebo1 <- rnorm((n/2),2,5)
changePlacebo2 <- rnorm((n/2),1,5)
placebo$time2 <- placebo$score - changePlacebo1 
placebo$time3 <- placebo$time2 - changePlacebo2 

both <- rbind(treatment, placebo)
both$subject <- as.factor(1:n)
both_long <- pivot_longer(both, cols = c('score','time2','time3'), names_to = 'time')
ggplot(both_long, aes(x=value, fill=time)) + facet_grid(~group) + geom_density(alpha=0.2) + theme_minimal()
```

So, the mean change of the treatment group in time 1 is `r mean(changeTrt1)` and in time 2 its `r mean(changeTrt2)`.
The mean changes in the placebo group are time1 = `r mean(changePlacebo1)` and time2 = `r mean(changePlacebo2)`
# Effect of statistical model
Now we can compare the effect of mixed level model or other on the ability to recover data

## Basic mixed level
We start with the basic mixed level design, subjects as random effect

```{r}
model1 <- lmer(value ~ time * group + (1|subject), data= both_long)
summary(model1)
anova(model1)
```

time as factor

```{r}
model1.1 <- lmer(value ~ as.factor(time) * group + (1|subject), data= both_long)
summary(model1.1)
anova(model1.1)
```

The model recovers the data well, but not perfect. It finds the general time effect, the group effect and the interaction.

## Mixed level with subjects nested within groups
```{r}
model2 <- lmer(value ~ time * group + (1|group/subject), data= both_long)
summary(model2)
anova(model2)
```

Model doesn not converge (?)

## Next - adding baseline 

```{r}
# first build the dataset
both2 <- both
both2$baseline <- both2$score
both2_long <- pivot_longer(both2, cols = c('score','time2','time3'), names_to = 'time')
```

### Using baseline and all levels
```{r}
model3 <- lmer(value ~ time * group + baseline + (1|subject), data= both2_long)
summary(model3)
#anova(model3)
```

time as factor
```{r}
model3.1 <- lmer(value ~ as.factor(time) * group + baseline + (1|subject), data= both2_long)
summary(model3.1)
#anova(model3)
```


### Using baseline instead of level 1 of the time variable

```{r}
both2_long2 <- dplyr::filter(both2_long, time!='score')
model4 <- lmer(value ~ time * group + baseline + (1|group/subject), data= both2_long2)
summary(model4)
anova(model4)
```

time as factor
```{r}
model4.1 <- lmer(value ~ as.factor(time) * group + baseline + (1|group/subject), data= both2_long2)
summary(model4.1)
anova(model4.1)

```

# Test correlation of the data

```{r}
p1 <- plot(treatment$score, treatment$time2)
cor.test(treatment$score, treatment$time2)
p2 <- plot(treatment$time2, treatment$time3)
cor.test(treatment$time2, treatment$time3)
```

## NLME

### CS
The Std.Error is the same at all timepoints (because it is "symmetric").
```{r}
LME_CS <- lme(value ~ as.factor(time) * group,
              random = ~ 1 | subject,  
              correlation=corCompSymm(form=~ 1 | subject),
              data=both2_long,
              na.action = "na.exclude")

summary(LME_CS)
```
### US
The Std.Error is different at different measurement time points.
```{r}
LME_US <- lme(value ~ as.factor(time) * group,
              random = ~ 1 | subject,  
              correlation=corSymm(form=~ 1 | subject),
              data=both2_long,
              na.action = "na.exclude")

summary(LME_US)
```
### AR(1)
The Std.Error is the same at all timepoints (because it is "symmetric").
```{r}
LME_AR <- lme(value ~ as.factor(time) * group,
              random = ~ 1 | subject,  
              correlation=corAR1(form=~ 1 | subject),
              data=both2_long,
              na.action = "na.exclude")

summary(LME_AR)
```

### Compare models
CS vs AR(1) vs US
```{r}
anova(LME_CS, LME_AR, LME_US)
```

AR shows best fit, but not sig diff.



### Check Assumptions of AR(1) model

[@zuur2010][@gelman2006]

Assumption 1: Linearity Because the DV is a Categorical variable, linearity is not assessed.

Assumption 2: Homogeneity of Variance

```{r Final Model FIX Assumptions 2, echo=FALSE, warning=TRUE, message=TRUE}
plot(resid(LME_AR))
```

Assumption 2: Normally distributed Residuals

```{r Final Model FIX Assumptions 3, echo=FALSE, warning=TRUE, message=TRUE}
qqnorm(resid(LME_AR))
```
## EMM for groups

```{r Final Model EMM, echo=FALSE, warning=TRUE, message=TRUE}
emmeans(LME_AR, ~ group*as.factor(time))
```

```{r Figure 2.2, message=TRUE, warning=TRUE, echo=FALSE}
## Extract adjusted means
Adjusted_means_fix <- as_tibble(emmeans(LME_AR, ~ group*time)) %>% 
  mutate(time = as.factor(time))

L1 <- ggplot(Adjusted_means_fix, aes(x=time, y = Outcome)) +
  geom_line(data = Adjusted_means_fix , aes(x = time, y = emmean, 
                                            group = group,
                                            color=group,
                                            linetype = group)) +
  geom_point(data = Adjusted_means_fix, aes(x = time, y = emmean, 
                                            group = group,
                                            color=group), 
             size = 1.5) +
  geom_errorbar(data = Adjusted_means_fix, aes(x = time, y = emmean, 
                                               ymin = lower.CL, ymax = upper.CL,
                                               color = group), 
                width = .1) +
  scale_y_continuous(expand = c(0,0),
                     limits = c(30, 65),
                     breaks = c(30,35,40, 45, 50, 55, 60, 65)) +
  scale_x_discrete(
    breaks = c("score","time2","time3"),
    label = c("Baseline", "T2", "T3")) +
  ylab("PCL Score") +
  xlab("Study Period") +
  theme_cowplot() +
  scale_color_manual(values=c("#0D0087FF", "#d11141"),
                     labels = c("placebo", "trt")) +
  theme(
    plot.title = element_text(size=11),
    axis.title.x = element_text(size=11, 
                                margin = margin(t = 5, r = 0, b = 0, l = 0), 
                                colour = "black"),
    axis.title.y = element_text(size=11, 
                                margin = margin(t = 0, r = 5, b = 0, l = 0), 
                                colour = "black"),
    axis.text.x = element_text(size=11, color = "black"),
    axis.text.y = element_text(size=11, color = "black", 
                               margin = margin(t = 0, r = 2, b = 0, l = 0)),
    axis.ticks = element_line(colour = "black", size=0.4),
    panel.grid.major.x = element_blank(), 
    panel.grid.major.y = element_line(size=.2, color="grey66"), 
    panel.grid.minor.y = element_blank(), 
    panel.grid.minor.x = element_blank(),
    legend.position = c(.70, .85),
    legend.title = element_blank(),
    legend.margin = margin(t = 3, r = 2, b = 3, l = 2),
    legend.box.background = element_rect(fill = "white", color="black", size=0.4),
    legend.box.margin = margin(t = 3, r = 2, b = 3, l = 2))
L1
```
