---
title: "Baseline Simulations"
author: "Or Duek & Tobias R. Spiller"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: yes
---

```{r}
# load libraries
require(ggplot2)
require(tidyr)
require(lme4)
require(lmerTest)
```

# Introduction
Here we will simulate randomized controlled trial of two groups (placebo and treatment) for 3 (or more) time points.
The main idea is to test the effect of using the baseline score as a covariate in a mixed level model.

# Generating Data
First, the total number of patients
```{r}
n = 100 # number of subjects
score <- rnorm(n, 55, 10) # generate data for PCL scores of PTSD patients
hist(score)
mean(score)
```

Now we shal divide them to two groups.
Group 1 (trt) will improve an average of 10 points at time 2 and another 5 at time 3
Group 2 (placebo) will improve an average of 2 points on time 2 and 1 on time 3

```{r}
scoreTrt <-  score[1:(n/2)]
scorePlacebo <- score[((n/2)+1):n]
treatment <- data.frame(score = scoreTrt)
placebo <- data.frame(score = scorePlacebo)
treatment$group <- 'trt'
placebo$group <- 'placebo'
# and combine into your new data frame vegLengths
both <- rbind(treatment, placebo)
ggplot(both, aes(x=score, fill = group)) + geom_density(alpha = 0.2) + theme_minimal()
```
So we see we start with similar distribution.
Now lets generate the change

```{r}
# in treatment group we remove an average of 10 points
changeTrt1 <- rnorm((n/2),10,5)
changeTrt2 <- rnorm((n/2),5,5)

treatment$time2 <- treatment$score - changeTrt1 
treatment$time3 <- treatment$time2 - changeTrt2 

# now for the plcebo group
changePlacebo1 <- rnorm((n/2),2,5)
changePlacebo2 <- rnorm((n/2),1,5)
placebo$time2 <- placebo$score - changePlacebo1 
placebo$time3 <- placebo$time2 - changePlacebo2 

both <- rbind(treatment, placebo)
both$subject <- as.factor(1:n)
both_long <- pivot_longer(both, cols = c('score','time2','time3'), names_to = 'time')
ggplot(both_long, aes(x=value, fill=time)) + facet_grid(~group) + geom_density(alpha=0.2) + theme_minimal()
```

So, the mean change of the treatment group in time 1 is `r mean(changeTrt1)` and in time 2 its `r mean(changeTrt2)`.
The mean changes in the placebo group are time1 = `r mean(changePlacebo1)` and time2 = `r mean(changePlacebo2)`
# Effect of statistical model
Now we can compare the effect of mixed level model or other on the ability to recover data

## Basic mixed level
We start with the basic mixed level design, subjects as random effect

```{r}
model1 <- lmer(value ~ time * group + (1|subject), data= both_long)
summary(model1)
anova(model1)
```

Time as factor

```{r}
model1.1 <- lmer(value ~ as.factor(time) * group + (1|subject), data= both_long)
summary(model1.1)
anova(model1.1)
```

The model recovers the data well, but not perfect. It finds the general time effect, the group effect and the interaction.

## Mixed level with subjects nested within groups
```{r}
model2 <- lmer(value ~ time * group + (1|group/subject), data= both_long)
summary(model2)
anova(model2)
```

Model doesn not converge (?)

## Next - adding baseline 

```{r}
# first build the dataset
both2 <- both
both2$baseline <- both2$score
both2_long <- pivot_longer(both2, cols = c('score','time2','time3'), names_to = 'time')
```

### Using baseline and all levels
```{r}
model3 <- lmer(value ~ time * group + baseline + (1|subject), data= both2_long)
summary(model3)
#anova(model3)
```

Time as factor
```{r}
model3.1 <- lmer(value ~ as.factor(time) * group + baseline + (1|subject), data= both2_long)
summary(model3.1)
#anova(model3)
```


### Using baseline instead of level 1 of the time variable

```{r}
both2_long2 <- dplyr::filter(both2_long, time!='score')
model4 <- lmer(value ~ time * group + baseline + (1|group/subject), data= both2_long2)
summary(model4)
anova(model4)
```

Time as factor
```{r}
model4.1 <- lmer(value ~ as.factor(time) * group + baseline + (1|group/subject), data= both2_long2)
summary(model4.1)
anova(model4.1)

```

# Test correlation of the data

```{r}
p1 <- plot(treatment$score, treatment$time2)
cor.test(treatment$score, treatment$time2)
p2 <- plot(treatment$time2, treatment$time3)
cor.test(treatment$time2, treatment$time3)
```


